---
title: "Time Series with R"
author: "Sara Geremia"
date: "2024-05-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, 
                      error = FALSE, message = FALSE)

```

Let's work with dates and times in R.

# Time Series

A time series is a sequence of data in chronological order. It is very common for any type of data to be recorded sequentially, or over time. And we find time series data everywhere, and especially in Financial and Economic applications. Examples include: monthly values of the House Price Index.


We will focus on the `lubridate` package, which makes it easy to work with dates and times in R.

```{r}
library(tidyverse)
library(lubridate)
```

## Create dates/times

You should always use the simplest possible data type that works for your needs:
avoid date-time data type if not necessary.
Date-times are substantially more complicated because of the need to handle time zones.

To get the current date or date-time you can use `today()` or `now()`:

```{r}
today()

now()
```

Date/time data often comes as strings. 
Identify the order in which year, month, and day appear in your dates, then use the 
appropriate functions:

```{r}
ymd("2024-05-24")
mdy("May 24th, 2024")
dmy("24-May-2024")
```

These functions also take unquoted numbers. 

```{r}
ymd(20240524)
```

Sometimes you’ll have the individual components of the date-time spread across multiple columns:

```{r}
# load data
# install.packages("nycflights13")
library(nycflights13)
```


```{r}
Sys.setlocale("LC_ALL", "en_US")

flights <- flights %>% 
  select(year, month, day, hour, minute) %>% 
  mutate(dep_time = make_datetime(year, month, day, hour, minute))
  
```

## Visualize dates/times

The most common first step when conducting time series analysis is to display your time series dataset in a visually intuitive format. 

The defining feature is that time is indexed on the horizontal axis, and the observations are shown from the first, on the left, to the last, on the right. A line is commonly added to connect neighboring observations, to improve interpretability and to emphasize any trends or patterns.

```{r}
flights %>% 
  ggplot(aes(dep_time)) + 
  geom_freqpoly(binwidth = 86400) # 86400 seconds = 1 day
```

```{r}
flights %>% 
  filter(dep_time < ymd("2013-01-02")) %>% 
  ggplot(aes(dep_time)) + 
  geom_freqpoly(binwidth = 600) # 600 s = 10 minutes
```

We can use wday() to see that more flights depart during the week than on the weekend:

```{r}
flights %>% 
  mutate(wday = wday(dep_time, label = TRUE)) %>% 
  ggplot(aes(x = wday)) +
    geom_bar()
```

Some time series data is exactly evenly spaced.
Some time series data is only approximately evenly spaced. 

The analysis of time series data proceeds with some simplifying assumptions: The first assumption is that consecutive observations are equally spaced. Secondly, a discrete-time observation index is applied. In practice, this may only hold approximately, and sometimes data may be missing. For example, daily log returns on a stock may only be available for weekdays, and data may not be available for certain holidays. Monthly House Price Index values are equally spaced by month, but not by days.

Sometimes there are missing values in time series data, denoted NA in R, and it is useful to know their locations. It is also important to know how missing values are handled by various R functions. Sometimes we may want to ignore any missingness, but other times we may wish to impute or estimate the missing values.

The mean() function calculates the sample mean, but it fails in the presence of any NA values. Use mean(___, na.rm = TRUE) to calculate the mean with all missing values removed. It is common to replace missing values with the mean of the observed values. 

```{r}

flights_dt <- flights %>% 
  count(year, month, day) %>% 
  mutate(dep_time = make_date(year, month, day)) %>% 
  filter(dep_time < ymd("2014-01-01"))


flights_dt_na <- flights_dt
flights_dt_na$n[flights_dt_na$dep_time > ymd("2013-01-31") & flights_dt_na$dep_time < ymd("2013-02-28")] <- mean(flights_dt_na$n, na.rm = TRUE)


```

```{r}
ggplot() +
  geom_line(data = flights_dt, aes(x = dep_time, y = n), color = "red", linetype = "dotted")+
  geom_line(data = flights_dt_na, aes(x = dep_time, y = n)) 
```

 Does this simple data imputation scheme appear adequate?
 
 Based on your plot, it seems that simple data imputation using the mean is not a great method to approximate what's really going on in the flights data.
 

## Create Time Series

 
Given a vector of numbers you can apply the ts() function to create a time series object. Such objects are of the `ts` class. They represent data that is at least approximately evenly spaced over time.
 
```{r}
data_vector <- c(10, 6, 11, 8, 10, 3, 6, 9)
print(data_vector)
plot(data_vector)

```

```{r}
time_series <- ts(data_vector)
```


When you plot the result using the plot() function the time index and label is automatically added to the horizontal axis. By default, R uses a simple observation index starting from 1 as the time index.

```{r}
print(time_series)
plot(time_series)
```

If you want the time series to start in the year 2001 with 1 observation per year you should apply the ts() function with the additional arguments start = 2001 and frequency = 1 as shown. Now when you plot the result you can see an updated time axis, running from 2001 through 2008.

```{r}
time_series <- ts(data_vector, start = 2001, frequency = 1)
plot(time_series)
```

A time series object is a vector (univariate) or matrix (multivariate) with additional attributes, including time indices for each observation, the sampling frequency and time increment between observations, and the cycle length for periodic data. 

The advantage of creating and working with time series objects of the ts class is that many methods are available for utilizing time series attributes, such as time index information.

Let's consider the eu_stocks dataset (available in R by default as EuStockMarkets). This dataset contains daily closing prices of major European stock indices from 1991-1998, specifically, from Germany (DAX), Switzerland (SMI), France (CAC), and the UK (FTSE). The data were observed when the markets were open, so there are no observations on weekends and holidays. We will proceed with the approximation that this dataset has evenly spaced observations and is a four dimensional time series.

 
```{r}
library(datasets)
data(EuStockMarkets)

```

```{r}
# Check whether EuStockMarkets is a ts object
is.ts(EuStockMarkets)

# View the start, end, and frequency of EuStockMarkets
start(EuStockMarkets)
end(EuStockMarkets)
frequency(EuStockMarkets)



# Generate a simple plot of EuStockMarkets
plot(EuStockMarkets)

# Use ts.plot with EuStockMarkets
ts.plot(EuStockMarkets, col = 1:4, xlab = "Year", ylab = "Index Value", main = "Major European Stock Indices, 1991-1998")

# Add a legend to your ts.plot
legend("topleft", colnames(EuStockMarkets), lty = 1, col = 1:4, bty = "n")

```

Let's try to make a better visualization.

```{r}
library(forecast)
autoplot(EuStockMarkets) + 
  theme_minimal() +
  labs(x = "Year", y = "Index Value", title = "Major European Stock Indices, 1991-1998") +
  scale_x_discrete(limits = seq(1991, 1998))
```


# Removing Trends

## Logarithm Transformations

The logarithmic function `log()` is a data transformation that can be applied to positively valued time series data. It slightly shrinks observations that are greater than one towards zero, while greatly shrinking very large observations. This property can stabilize variability when a series exhibits increasing variability over time. It may also be used to linearize a rapid growth pattern over time.

```{r}
rapid_growth <- ts(c(
  505.9547, 447.3556, 542.5831, 516.0634, 506.9599, 535.0162, 496.9291,
  497.5626, 577.2483, 536.8560, 541.2459, 473.4978, 550.9890, 569.4106,
  522.9152, 487.2002, 594.6108, 591.1740, 615.9868, 621.3175, 607.1250,
  587.0367, 554.1554, 644.1172, 509.7000, 607.0943, 603.5512, 613.6216,
  544.9143, 670.8118, 687.1316, 615.5817, 711.1873, 694.2979, 681.9293,
  659.1403, 642.7021, 601.5301, 666.7623, 650.9657, 606.0913, 696.6788,
  641.6025, 855.7719, 667.3291, 573.4914, 791.7333, 751.5914, 610.7948,
  624.6503, 833.2990, 639.8867, 736.8283, 772.2923, 686.8865, 667.7631,
  712.9415, 918.1838, 656.1089, 700.4972, 683.4933, 781.7380, 715.6843,
  808.2875, 820.7795, 656.8856, 733.3400, 773.5387, 641.2027, 932.2119,
  680.6766, 988.2828, 664.8986, 813.5283, 883.4088, 924.2749, 969.4321,
  777.3293, 880.9984, 971.3583, 902.9584, 1020.7457, 1075.1483, 886.1707,
  889.6322, 950.3908, 878.0395, 1043.7676, 901.1090, 1079.6584, 933.9054,
  921.9433, 870.8071, 811.1398, 1004.2677, 1008.1758, 1189.4893, 751.9706,
  947.4753, 886.5153, 1074.8943, 1101.1307, 1130.1855, 975.8495, 948.1610,
  1177.8227, 1227.1271, 976.9957, 836.7089, 1323.6047, 852.3532, 1200.8262,
  1274.4788, 1349.2614, 1102.6334, 1324.8566, 1268.7187, 1058.2289, 1204.0872,
  1084.6503, 1284.4305, 1195.2843, 1058.4262, 1188.0577, 1166.5934, 1064.6946,
  1429.0685, 1070.8528, 1539.3305, 1467.1571, 1127.7058, 1296.0717, 1555.2741,
  1332.9037, 1315.4236, 1189.2462, 1482.4339, 1240.9287, 1237.7720, 1468.6083,
  1328.5457, 1589.5078, 1373.1630, 1503.5563, 1659.9376, 1704.6137, 1550.4638,
  1625.8026, 1873.8582, 1370.6209, 1439.7114, 1447.4369, 1579.9158, 1681.2571,
  1661.6059, 1311.8468, 1326.0308, 1323.0995, 1550.4863, 1606.2042, 1768.5401,
  1509.8368, 1592.1086, 1627.6188, 1544.6329, 1439.5234, 1682.3518, 1850.7097,
  1673.3801, 1832.4272, 1672.2672, 1781.5768, 1659.2899, 1970.0389, 2044.7124,
  1929.0902, 1891.7042, 1487.1577, 2013.8722, 1796.7886, 1977.0183, 1516.9552,
  1650.6039, 1523.2834, 1696.6181, 1627.2609, 1787.2968, 1567.2874, 1881.9963,
  2318.9833, 1941.9879, 1820.2797, 2154.8123, 2261.5471, 2052.2214, 2079.1710,
  2010.0609, 2145.2606, 1775.3008, 2013.4070
))
```

```{r}
autoplot(rapid_growth) + 
  theme_minimal() +
  labs(x = "Year", y = "Index Value") 
```

Apply the log() function to rapid_growth

```{r}
linear_growth <- log(rapid_growth)

autoplot(linear_growth) + 
  theme_minimal() +
  labs(x = "Year", y = "Index Value") 
```

The logarithmic transformation helps stabilize your data by inducing linear growth over time.

## First Difference Transformations

The first difference transformation of a time series $Y_t$
 consists of the differences (changes) between successive observations over time, that is 
$Y_t - Y_{t-1}$.

Differencing a time series can remove a time trend. The function `diff()` will calculate the first difference or change series. A difference series lets you examine the increments or changes in a given time series. It always has one fewer observations than the original series.

```{r}
dz <- diff(linear_growth)

autoplot(dz) + 
  theme_minimal() +
  labs(x = "Time", y = "Index Value") 
```

By removing the long-term time trend, you can view the amount of change from one observation to the next.

## $s$ Difference Transformations

For time series exhibiting seasonal trends, seasonal differencing can be applied to remove these periodic patterns. For example, monthly data may exhibit a strong twelve month pattern. In such situations, changes in behavior from year to year may be of more interest than changes from month to month, which may largely follow the overall seasonal pattern.

```{r}
y <- ts(c(
  -4.198033, 9.569009, 5.175143, -9.691646, -3.215294, 10.843669,
  6.452159, -10.833559, -2.235351, 10.119833, 6.579646, -8.656565,
  -2.515001, 9.837434, 7.386194, -8.243504, -4.264033, 8.898861,
  8.544336, -8.066913, -4.023025, 9.822679, 7.772852, -6.587777,
  -3.459171, 10.613851, 7.374450, -5.798715, -1.204711, 11.429236,
  7.570047, -4.968384, -2.003787, 11.941348, 9.406672, -4.396585,
  -1.555579, 12.599877, 8.502916, -3.728968, -2.827000, 13.375981,
  8.128941, -3.149249, -2.799473, 13.710570, 6.755217, -3.779744,
  -3.768274, 13.625336, 6.537931, -3.249098, -5.024191, 13.355373,
  6.931161, -3.527354, -5.197329, 11.579791, 7.162449, -1.894607,
  -5.777797, 12.482695, 6.208088, -3.434038, -7.080721, 11.413656,
  6.741990, -3.532376, -8.393542, 12.507261, 6.473175, -3.745246,
  -9.426209, 12.380817, 8.048243, -2.831528, -7.301893, 12.765838,
  8.223699, -4.448131, -6.963558, 12.034005, 7.574925, -5.402218,
  -6.568198, 10.896482, 7.276571, -4.037873, -6.723013, 12.180815,
  8.285162, -4.159342, -6.360670, 12.753018, 8.665912, -5.440538,
  -4.874932, 12.600197, 8.162589, -6.539572
))

autoplot(y) + 
  theme_minimal() +
  labs(x = "Time", y = "Index Value") 
```
The function `diff(..., lag = s)` will calculate the lag s difference or length s seasonal change series. For monthly or quarterly data, an appropriate value of s would be 12 or 4, respectively. The `diff()` function has `lag = 1` as its default for first differencing. Similar to before, a seasonally differenced series will have $s$ fewer observations than the original series.

```{r}
dy <- diff(y, lag = 4)

autoplot(dy) + 
  theme_minimal() +
  labs(x = "Time", y = "Index Value") 
```

Once again differencing allows you to remove the longer-term time trend - in this case, seasonal volatility - and focus on the change from one period to another.

# White Noise (W-N) Time Series

The white noise (WN) model is a basic time series model. It is also a basis for the more elaborate models we will consider. We will focus on the simplest form of WN, independent and identically distributed data.

The `arima.sim()` function can be used to simulate data from a variety of time series models. For now we note that the `ARIMA(0, 0, 0)` model is simply the WN model.

```{r}
# Simulate a WN model with list(order = c(0, 0, 0))
white_noise <- arima.sim(model = list(order = c(0, 0, 0)), n = 100)

autoplot(white_noise) + 
  theme_minimal() +
  geom_hline(yintercept = 0, color = "red", linetype = "dotted")  +
  labs(x = "Time", y = "Index Value") 
```

```{r}
# Simulate from the WN model with: mean = 100, sd = 10
white_noise_2 <- arima.sim(list(order = c(0, 0, 0)), n = 100, mean = 100, sd = 10)

autoplot(white_noise_2) + 
  theme_minimal() +
  geom_hline(yintercept = 100, color = "red", linetype = "dotted")  +
  labs(x = "Time", y = "Index Value") 
```

For a given time series $Y$ we can fit the white noise (WN) model using the `arima(..., order = c(0, 0, 0))` function. Recall that the WN model is an `ARIMA(0,0,0)` model. Applying the `arima()` function returns information or output about the estimated model. For the WN model this includes the estimated mean, labeled `intercept`, and the estimated variance, labeled `sigma^2`.

```{r}
y_ts <- ts(c(
  109.76134, 98.31610, 100.63295, 88.74340, 101.87238, 104.62836, 96.64462,
  102.86194, 112.76247, 82.23219, 88.94434, 94.60318, 105.61113, 113.82776,
  104.82319, 101.29629, 82.98459, 88.96058, 100.60046, 91.03525, 100.19286,
  95.32537, 94.58643, 121.35830, 87.37874, 96.89007, 90.51310, 99.90843,
  102.56934, 104.96144, 104.73464, 88.00465, 100.77943, 121.64776, 85.77083,
  79.15428, 98.18847, 99.90006, 98.91084, 101.64422, 102.79526, 84.65112,
  96.47870, 105.81547, 98.51869, 105.24366, 109.61264, 85.14201, 82.81442,
  103.03629, 93.56967, 98.08922, 81.25461, 109.18554, 80.43181, 103.55953,
  80.22269, 84.97477, 107.78363, 92.61288, 99.80293, 107.25085, 98.66378,
  91.92275, 98.32642, 112.73402, 96.02811, 92.64909, 83.08484, 97.22196,
  106.61361, 97.36943, 108.78465, 104.91858, 84.44343, 85.60786, 96.51529,
  94.18105, 85.02851, 63.26622, 87.22137, 103.52295, 105.60216, 103.26039,
  101.11519, 108.45697, 97.67631, 103.01081, 100.61756, 105.58108, 98.72722,
  98.43996, 90.78219, 92.74599, 102.93762, 83.95306, 110.15937, 104.17578,
  99.27876, 103.25115
))
```

```{r}
# Fit the WN model to y_ts using the arima command
arima(y_ts, order = c(0, 0, 0))

# Calculate the sample mean and sample variance of y_ts

mean(y_ts)
var(y_ts)
```

From the comparisons you can see that the arima() function estimates are very close to the sample mean and variance estimates, in fact identical for the mean.

# Random Walk (R-W) Time Series

The random walk (RW) model is also a basic time series model. It is the cumulative sum (or integration) of a mean zero white noise (WN) series, such that the first difference series of a RW is a WN series. Note for reference that the RW model is an `ARIMA(0, 1, 0)` model, in which the middle entry of 1 indicates that the model's order of integration is 1.

The arima.sim() function can be used to simulate data from the RW by including the model = list(order = c(0, 1, 0)) argument. We also need to specify a series length n. Finally, you can specify a sd for the series (increments), where the default value is 1.

```{r}
set.seed(123)
# Generate a RW model using arima.sim
random_walk <- arima.sim(model = list(order = c(0, 1, 0)), n = 100)

autoplot(random_walk) + 
  theme_minimal() +
  labs(x = "Time", y = "Index Value") 

```

```{r}
# Calculate the first difference series
random_walk_diff <- diff(random_walk)

autoplot(random_walk_diff) + 
  theme_minimal() +
  labs(x = "Time", y = "Index Value") 

```

The first difference of your random_walk data is white noise data. This is because a random walk is simply recursive white noise data. By removing the long-term trend, you end up with simple white noise.


A random walk (RW) need not wander about zero, it can have an upward or downward trajectory, i.e., a drift or time trend. This is done by including an intercept in the RW model, which corresponds to the slope of the RW time trend.

To simulate data from the RW model with a drift you again use the `arima.sim()` function with the `model = list(order = c(0, 1, 0))` argument. This time, you should add the additional argument `mean = ...` to specify the drift variable, or the intercept.

```{r}
set.seed(123)
# Generate a RW model with a drift uing arima.sim
rw_drift <- arima.sim(model = list(order = c(0, 1, 0)), n = 100, mean = 1)

autoplot(rw_drift) + 
  theme_minimal() +
  labs(x = "Time", y = "Index Value") 

```

```{r}
# Now fit the WN model to the differenced data
model_wn <- arima(random_walk_diff, order = c(0, 0, 0))

# Store the value of the estimated time trend (intercept)
int_wn <- model_wn$coef

# Plot the original random_walk data
autoplot(random_walk) + 
  theme_minimal() +
  labs(x = "Time", y = "Index Value") +
  # Use geom_abline() to add time trend to the figure
  geom_abline(slope = int_wn, intercept = 0, 
              color = "red", linetype = "dotted")

```

