---
title: "Frequency Distributions and Visualization with R"
author: "Sara Geremia"
date: "2024-03-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
```

Load data

```{r}
load("practice/01-Descriptive/data/milan_egypop.RData")
```

```{r}
head(milan_egypop)
tail(milan_egypop)
```
The `str` function compactly display the internal structure of an R object, in this case a data frame.

```{r}
str(milan_egypop)
```


```{r}
names(milan_egypop) <- c("year", "id_nil", "nil", "age", "gender", "nat", "inhab")
```


If you want to count the relative frequency of certain observations in your sample, you can identify the interesting observations by using a logical expression; then calculate the fraction of observations it identifies.

## Example: Relative frequency distribution of Egyptians by gender in Milan in 2023

- **Base-R**

Compute the relative frequency of males and females in 2023

```{r}
# filter by gender and year
egypop23m <- milan_egypop[milan_egypop$gender == "Maschi" & milan_egypop$year == 2023, ]

# total number of inhabitants in 2023
N_egypop23 <- sum(milan_egypop$inhab[milan_egypop$year == 2023])

# absolute frequency of males in 2023
n_egypop23m <- sum(egypop23m$inhab)
# relative frequency of males in 2023
f_egypop23m <- n_egypop23m / N_egypop23
f_egypop23m


# relative frequency of females in 2023
1 - f_egypop23m

egypop23f <- milan_egypop[milan_egypop$gender == "Femmine" & milan_egypop$year == 2023, ]

n_egypop23f <- sum(egypop23f$inhab)
f_egypop23f <- n_egypop23f / N_egypop23
f_egypop23f
```

Visualize the frequency distribution: pie chart

```{r}
egypop23_gender <- c(rep("Maschi", n_egypop23m), rep("Femmine",  n_egypop23f))

table(egypop23_gender)
pie(table(egypop23_gender))

```

- **Tidyverse**

Compute the relative frequency of males and females in 2023

```{r}
egypop23_gender <- milan_egypop %>% 
  # filter by year
  filter(year == 2023) %>%
  # total number of inhabitants in 2023
  mutate(N = sum(inhab)) %>% 
  # perform operations by gender (separately on males and females)
  group_by(gender) %>% 
  # absolute and relative frequency by gender
  mutate(n = sum(inhab),
         f = n / N) %>%  
  # select useful variables 
  select(year, gender, n, f) %>% 
  # keep only distinct rows
  distinct()
```
### Pie Chart

```{r}
ggplot(egypop23_gender, aes(x="", y=f, fill=gender)) +
  geom_bar(stat="identity", width=1) +
  coord_polar("y", start=0)
```
### Bar Plot

```{r}
ggplot(egypop23_gender, aes(x=gender, y=f, fill=gender)) +
  geom_col(width=.8)
```

Exercise 1: Compute the relative frequency distribution of Egyptians by age in Milan in 2023.
Is it easy to do it in base-R?

## Example: Absolute and relative frequency distribution of Egyptians by year and gender in Milan


- **Tidyverse**

Compute the relative frequency of males and females across years

```{r}
egypop_gender <- milan_egypop %>% 
  # total number of inhabitants by gender
  group_by(gender) %>% 
  mutate(N = sum(inhab)) %>% 
  # perform operations by gender and year
  group_by(gender, year) %>% 
  # absolute frequency by gender
  mutate(n = sum(inhab),
         f = n / N) %>%  
  # select useful variables 
  select(year, gender, n, f) %>% 
  # keep only distinct rows
  distinct()
```
### Bar Plot

```{r}
ggplot(egypop_gender, aes(x=year, y=n, fill=gender)) +
  geom_col(width=.8) +
  facet_grid(cols = vars(gender))

ggplot(egypop_gender, aes(x=year, y=f, fill=gender)) +
  geom_col(width=.8) +
  facet_grid(cols = vars(gender))
```

### Trend plot

```{r}
ggplot(egypop_gender, aes(x=year, y=f, color=gender, group=gender)) +
  geom_point() +
  geom_line()
```

Execise 2: Compute the absolute and relative frequency distribution of Egyptians by age and gender in Milan